---
title: "NRM Tutorial"
author: "Giona Casiraghi"
date: "7/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the tutorial companion to the article ``Multiplex Network Regression: How do relations drive interactions?''.
It provides the code to reproduce the results shown in the application section of the article.
The source code of this tutorial is available at [https://github.com/gi0na/nrm-tutorial](https://github.com/gi0na/nrm-tutorial).

## Installation

The R package `ghypernet` provides the routines used to estimate network regression models.
Install the latest version from GitHub, to ensure to have all latest features.
The R package `adjConvertR` provides some useful and fast functions to construct adjacency matrices from dataframes.

```{r installation, echo = T, results = 'hide'}
# install ghypernet
devtools::install_github('gi0na/ghypernet')
# install adjConvertR
devtools::install_github('gi0na/adjConvertR')

library(ghypernet)
library(adjConvertR)
library(readr)
```

## Load SocioPatterns Data

The next step is to download the data from \url{sociopatterns.org}, and load it into the rsession.

```{r data}
# create data directory
dir.create(file.path('data'), showWarnings = FALSE)

# download the files
download.file(
  url = 'http://www.sociopatterns.org/wp-content/uploads/2015/07/High-School_data_2013.csv.gz',
  destfile = 'data/High-School_data_2013.csv.gz')
download.file(
  url = 'http://www.sociopatterns.org/wp-content/uploads/2015/09/metadata_2013.txt',
  destfile = 'data/metadata_2013.txt')
download.file(
  url = 'http://www.sociopatterns.org/wp-content/uploads/2015/07/Facebook-known-pairs_data_2013.csv.gz',
  destfile = 'data/Facebook-known-pairs_data_2013.csv.gz')
download.file(
  url = 'http://www.sociopatterns.org/wp-content/uploads/2015/07/Friendship-network_data_2013.csv.gz',
  destfile = 'data/Friendship-network_data_2013.csv.gz')

# decompress
files <- list.files(path = "data/", pattern = '*.gz')
for(f in files){
  R.utils::gunzip(paste0('data/',f), overwrite=TRUE)
}

# load
# contact edge list
el_contacts <- read_table2("~/Documents/Matematica/SG/data/SocioPatterns/highschool/High-School_data_2013.csv", 
                           col_names = FALSE, col_types = cols(X2 = col_character(), X3 = col_character()))

# vertices metadata
metadata <- read_table2("~/Documents/Matematica/SG/data/SocioPatterns/highschool/metadata_2013.txt", 
                        col_names = FALSE, col_types = cols(X1 = col_character()))

# facebook edge list
el_fb <- read_table2("~/Documents/Matematica/SG/data/SocioPatterns/highschool/Facebook-known-pairs_data_2013.csv", 
                        col_names = FALSE, col_types = cols(X1 = col_character(), X2 = col_character(), X3 = col_logical()))

# friendship edge list
el_fr <- read_table2("~/Documents/Matematica/SG/data/SocioPatterns/highschool/Friendship-network_data_2013.csv", 
                     col_names = FALSE, col_types = cols(X1 = col_character(), X2 = col_character()))
```

Create the adjacency matrices for all the layer of the network regression model.

```{r adjacencies preparation}
# adjacency matrix
adj <- get_adjacency(x = el_contacts, select_cols = 2:3, multiedge = TRUE, sparse = FALSE, directed = F)

# metadata ordering and removing missing vertices
metadata <- metadata[-which(!(metadata$X1 %in% rownames(adj))),]
maporder <- function(id,v) which(id==v)
metadata <- metadata[Vectorize(FUN = maporder, vectorize.args = 'id')(id=rownames(adj),metadata$X1),]

# fb adjacency matrix
fb <- get_adjacency(x = el_fb, select_cols = 1:3, multiedge = FALSE, sparse = FALSE, directed = T, nodes = rownames(adj))

# friendship net adjacency matrix
fr <- get_adjacency(x = el_fr, sparse = FALSE, directed = T, nodes = rownames(adj))

# check that names of different matrices match
all(rownames(adj)==rownames(fb))
all(rownames(adj)==rownames(fr))
```
Construct the predictors for the `friendship' layers:
```{r friendship}
# construct predictor for reciprocal friendships
fr_pred <- fr*t(fr)
fr_pred[fr_pred==1] <- exp(1)
fr_pred[fr_pred==0] <- 1

# construct predictor for non-reciprocated friendships
fr_prednr <- fr+t(fr)
fr_prednr[fr_prednr==1] <- exp(1)
fr_prednr[fr_prednr<exp(1)] <- 1

frlist <- list(fr=fr_pred, fr_half=fr_prednr)
```

Construct the predictors for the `facebook' layers:
```{r facebook}
# construct predictor for facebook connections
fb_pred <- fb*(exp(1)-1)+1

# construct dummy variable for missing connections
indat <- unique(c(el_fb$X1,el_fb$X2))
fb_eps <- matrix(1,nrow(adj), ncol(adj))
fb_eps[
  -as.integer(plyr::mapvalues(x = indat, from = rownames(fb), to = seq_along(rownames(fb)))),
  -as.integer(plyr::mapvalues(x = indat, from = rownames(fb), to = seq_along(rownames(fb))))
  ] <- exp(1)

fblist <- list(fb=fb_pred, fb_eps=fb_eps)
```

## Network Regression Models
Estimate the nrm models using (a) the friendship social network as covariate, and (b) the facebook connections.
The function `nrm` in `ghypernet` takes argument: a list of the covariates as adjacency matrices, the adjacency matrix of the dependent variable, the directionality of the network and the presence of selfloops, the (optional) initial value to start the numerical optimisation.
Check the help of `nrm` with `?nrm`.

The first model we estimate is (a):
```{r (a)}
fr.m <- nrm(w = frlist, adj = adj, directed = FALSE, selfloops = FALSE, init=c(3,2))
fr.m
```

Model (b) is estimated in a similar fashion:
```{r (b)}
fb.m <- nrm(w = fblist, adj = adj, directed = F, selfloops = F, init=c(3,2))
fb.m
```
Note that model (a) explain much better than data compared to (b), as can be seen from McFadden $\rho^2$.

## Introducing the control variables

The next step is constructing the control variables using the metadata available about the nodes.
To do so, we first use a function that fills a matrix with Euler's number $e$ if two nodes have the same label, and $1$ otherwise.
```{r helper function}
samelabel <- function(v,w) ifelse((v==w), exp(1), 1)
samelabelv <- Vectorize(FUN = Vectorize(FUN = samelabel, vectorize.args = 'v'), vectorize.args = 'w')
```

Using the vectorized function `samelabelv` we can then construct the three control layers of our regression model.
```{r control layers}
class <- list(class=samelabelv(metadata$X2,metadata$X2))

topics <- substr(metadata$X2, start = 1, stop = 2)
topic <- list(topic=samelabelv(topics,topics))

samelabel <- function(v,w) ifelse((v==w & v!='Unknown'), exp(1), 1)
gender <- list(gender=samelabelv(metadata$X3,metadata$X3))
```

We can now introduce the control variables into the model estimated above.

```{r control-models}
(frcontrol.m <- nrm(w = c(class, topic, gender, frlist), adj = adj, directed = FALSE, selfloops = FALSE, init=c(4,2, .5, coef(fr.m))))
(fbcontrol.m <- nrm(w = c(class, topic, gender, fblist), adj = adj, directed = FALSE, selfloops = FALSE, init=c(4,2, .5, coef(fb.m))))
```

## Full model and model comparisons

The final step is the construction of the full model, using all 7 covariate layers.
Once we have estimated the model, we can compare the different models using AIC.

```{r full-model}
(full.m <- nrm(w = c(class, topic, gender, frlist,fblist), adj = adj, directed = FALSE, selfloops = FALSE, init=c(coef(frcontrol.m), coef(fb.m))))
AIC(fr.m,fb.m,frcontrol.m,fbcontrol.m,full.m)
```

Here we see that the largest contribution, among model (a) and (b), comes from (a).
That means, most of the explaining power for the observed interactions has to be ascribed to the friendship social network.
On the other hand, the explaining power provided by the facebook network vanishes as soon as the friendship network is added to the model.

Finally, we can use the `screenreg` function in the `texreg` package to nicely compare the different models and visualise them side by side.
First, though we need to create a method for `screenreg` that is compatible with `nrm` objects.
This will be included in the next version of `ghypernet`.

```{r texreg}
library(texreg)
## texreg package
## Texreg: does not (yet) support nrm or gyhpe-class
# use the extract()-function to make this available
extract.nrm.cluster <- function(model, ...){
  # calculate SE, tvalues and pvalues
  coeffic <- as.numeric(model$coef)
  stderr <- model$confint[,3]
  tvalues = abs(coeffic/stderr)
  pval <- 2 * stats::pnorm(-tvalues)
  
  # then create and return a texreg object (replace NULL with actual values):
  tr <- createTexreg(
    coef.names = names(model$coef),    # character vector of coefficient labels
    coef = coeffic,          # numeric vector with coefficients
    se = stderr,            # numeric vector with standard error values
    pvalues = pval,       # numeric vector with p-values
    gof.names = c("AIC", "McFadden $R^2$"),     # character vector with goodness-of-fit labels
    gof = c(model$AIC, model$R2)           # numeric vector of goodness-of-fit statistics
    #gof.decimal = NULL    # logical vector: GOF statistic has decimal points?
  )
  return(tr)
}
setMethod("extract", signature = className("nrm", "ghype"), 
          definition = extract.nrm.cluster)
```

```{r model table, results='asis'}
htmlreg(list(fr.m,fb.m,frcontrol.m,fbcontrol.m,full.m), digits = 3, 
       custom.coef.map =   list('class'='$R^{(C)}$', 
                                'fr'='$R^{(f)}$',
                                'topic'='$R^{(T)}$',
                                'fb'='$R^{(\\text{fb})}$',
                                'fb_eps'='$R^{(\\varepsilon)}$', 
                                'gender'='$R^{(G)}$', 
                                'fr_half'='$R^{(1/2f)}$'
       ),
       single.row = TRUE, doctype = FALSE,
       reorder.coef = c(c(1,3,6), c(2,7), 4:5), 
       groups = list('Control' = 1:3,  'Friendship' = 4:5,
                     'Facebook' = 6:7), 
       custom.model.names = c('$\\Theta^{(a)}$' , '$\\Theta^{(b)}$' , '$\\Theta^{(a\\dagger)}$' , '$\\Theta^{(b\\dagger)}$' , '$\\Theta^{(a\\dagger+b\\dagger)}$'))
```

## Stepwise Model Selection

Finally, we show how to perform a forward stepwise model selection using nrm.
The routine `nrm_selection` allows to do this automatically.
First, we need to construct the predictors as a nested list so that the algorithm can cycle through all of them.
```{r stepwise-preds}
# first, create list of predictors
preds <- c(class, 
           topic, 
           gender,
           frlist,
           # because we cannot split the fb layer from the dummy for the missing data, we nest them into another list
           fb=list(fblist)
           )

# then we wrap all predictors into lists using the function below
preds <- createPredictors(preds)

# finally, we need to fix the fb layer so that is only one list deep
preds$fb <- preds$fb$fb

# we perform a similar job with the initial values, to speed up the computation
init <- c(coef(frcontrol.m), coef(fb.m))
init[6] <- list(init[6:7])
init[7] <- NULL
```
Next, we can run the stepwise model selection routine.
```{r stepwise}
summary(selmodels <- nrm_selection(adj = adj, predictors = preds, directed = FALSE, selfloops = FALSE, init=init))
```

Finally, we can exploit again `texreg` for a nicer and clearer visualisation of the results.
To improve the readability, we rescale the AIC values such that the best model has AIC$=0$.
This is always possible as the AIC value is only useful when compared relative to another AIC value.
```{r aic-fix}
for(i in seq_along(selmodels$models)){
  selmodels$models[[i]]$AIC <- selmodels$models[[i]]$AIC-selmodels$models[[length(selmodels$models)]]$AIC
}
```

We use a similar snippet as the one above to print the results.
```{r stepwise-table, results='asis'}
htmlreg(selmodels$models, digits = 3, 
       custom.coef.map =   list('class'='$R^{(C)}$', 
                                'fr'='$R^{(f)}$',
                                'topic'='$R^{(T)}$',
                                'fb'='$R^{(\\text{fb})}$',
                                'fb_eps'='$R^{(\\varepsilon)}$', 
                                'gender'='$R^{(G)}$', 
                                'fr_half'='$R^{(1/2f)}$'
       ),
       single.row = TRUE, doctype = FALSE,
       reorder.coef = c(c(1,3,6), c(2,7), 4:5),
       groups = list('Control' = 1:3,  'Friendship' = 4:5,
                     'Facebook' = 6:7), 
       custom.model.names = c('(1)', '(2)', '(3)', '(4)','(5)','(6)'))
```